Untitled

### 2021-08-30 README.Rmd
add sections to manuscript
fix figure captions and plot colours

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

>>>>

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) <!-- DOI badge: [![DOI](.svg)](repo link) -->

<!-- Document structure provided by rrtools package -->

### 2021-08-30 analysis/paper/fig_tab.Rmd
further development of manuscript
fix figure captions and plot colours

<!-- ### Tables ### -->

<!-- ### Figures ### -->

@@

<!-- change colour of medium starches in D to match C -->
```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

>>>>

## Tables
<!-- ### Tables ### -->

```{r sample-perc}
# convert proportions to percent labels
perc_comb <- prop_comb %>% 
  mutate(across(c(s, m, l, total), scales::percent, 0.001))

perc <- prop %>%
  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
  add_row(tibble_row(perc_comb), .before = 3)
```

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorported into the samples.")
```

## Figures
<!-- ### Figures ### -->

@@

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```

### 2021-08-30 analysis/paper/paper.Rmd
fix figure captions and plot colours

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: |
  keyword 1; keyword 2
---

@@

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the rate of $\alpha$-amylase differs both between and within
individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

@@

<!-- INSERT PICTURES OF STARCHES -->

@@

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- change colour of medium starches in D to match C -->
```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```
<!-- Combine to a single figure? -->

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```
<!-- with Z-score standardisation -->

@@

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting, as many of the variables can be adjusted. 
The addition of known starch species in
separate samples also removes the identification bias against the smaller
undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring between individuals, and difficulties in accurately documenting 
the entirety of starch granules entering the oral cavity at any given moment,
especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.
It can also facilitate training of students and researchers on methods of 
dental calculus analysis, such as starch and phytolith extraction and 
identification, where it can replace the use of finite archaeological resources.  
<!-- we still need to replicate results and do more science... but pretty cool, right? -->

>>>>

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: |
  keyword 1; keyword 2
---

@@

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the rate of $\alpha$-amylase differs both between and within
individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

@@

<!-- INSERT MICROSCOPE PICTURES OF STARCHES -->

@@

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- Combine to a single figure? -->

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```

<!-- with Z-score standardisation -->

@@

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting, as many of the variables can be adjusted. 
The addition of known starch species in
separate samples also removes the identification bias against the smaller
undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring between individuals, and difficulties in accurately documenting 
the entirety of starch granules entering the oral cavity at any given moment,
especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, salivary flow, tooth position and morphology 
[@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
as well as evolutionary differences through time [@yatesOralMicrobiome2021].
It can also facilitate training of students and researchers on methods of 
dental calculus analysis, such as starch and phytolith extraction and 
identification, where it can replace the use of finite archaeological resources.  
<!-- we still need to replicate results and do more science... but pretty cool, right? -->
